Formalizing Commutative Algebra in Coq: \\ Nakayama's Lemma\thanks{Source code for this work is available on the following site: \url{https://github.com/ku-sldg/algebra}}

### 2022-12-11 nakayama/paper.tex
First edits on Nakayama paper

The importance of quotients in commutative algebra motivates our use of
equivalence relations to define the components of a group structure. If one
were to instead use the regular Leibniz equality, it would be difficult to
identify a quotient group and another group. However, by  defining a group in
terms of an arbitrary equivalence relation, in our theory a quotient group is 
simply defined as a group, but under an equivalence relation that is not the usual 
equality. 
Not much is lost due to Coq's rewrite tactics for setoids--types equipped with an equivalence relation---
can still be called upon.

@@

This argument has potentially infinitely many steps, and chose to avoid this
issue by including an axiom that in any non-unit \(x\) of a ring is contained in some
maximal ideal.

@@

\begin{minted}{coq}
Theorem nakayama:
  forall {n: nat}(basis: t M n),
    finitely_generated Mequiv Madd Mzero action basis ->
  (forall a: M, ideal_module_pred a) ->
  forall a: M, a =M= Mzero.
\end{minted}

To use as a building block in the formal proof of Nakayama's Lemma, we first formally stated and proved a lemma that gives a concrete description of the elements of the submodule $\mathfrak{m} M$  appearing in the statement of the the theorem.  
This lemma applies more generally, to any finitely generated module $M$ over a (not necessarily local) commutative ring, and any submodule of the form $IM$, for $I$ an arbitrary ideal. 
By definition, $IM$ consists of scalar combinations of elements of $M$ with coefficients in $I$. 
The lemma states that the elements from $M$ appearing in the expression can be chosen to be from any 
 fixed finite generating set for $M$. 
In other words, given any generating set  $u_{1},\dots, u_{n} \in M$  of a  finitely generated $R$-module $M$, every element $x$ of $I M$ can be written as a scalar combination $a_1 \cdot u_1 + a_2  \cdot u_2 + \cdots + a_n  \cdot u_n$, where each $a_i \in I$.

@@

The proof follows from a straightforward argument based on definitions, by induction on the number of elements in a fixed generating set for $M$, which we informally describe:  By definition, $x \in I M$ can  be written, for some positive integer $k$ and elements $r_i \in R$ and $w_i \in M$,  $1 \leq i \leq k$, as   $x = r_1 \cdot w_1 + r_2 \cdot w_2 + \cdots + r_k \cdot w_k$.
What's more, by definition of a finite generating set $u_{1},\dots, u_{n}$ for $M$,  each $w_i$ equals 
$c_{i1} \cdot u_1 + c_{i2} \cdot u_2 + \cdots + c_{in} \cdot u_n$ for appropriate choices of $c_{ij} \in R$. 
Hence, inductively applying associativity, 
\begin{align*}
  x = \sum_{i=1}^{k} r_i  \cdot  w_i  =  \sum_{i=1}^{k} r_i  \cdot  \left(  \sum_{j=1}^n c_{ij}  \cdot  u_j \right) 
  =   \sum_{j=1}^n  \sum_{i=1}^{k} r_i   \cdot  (c_{ij}  \cdot  u_j ) 
=  \sum_{j=1}^n \left( \sum_{i=1}^{k}  c_{ij}  \cdot   r_i \right)   \cdot  u_j.   
\end{align*}
By absorption in ideals, each  $c_{ij}  \cdot r_i$ is in $I$, and since ideals are closed under addition, we inductively conclude that 
the coefficient $a_j := c_{1j} \cdot  r_1 +  c_{2j} \cdot  r_2 +  \cdots c_{kj}\cdot r_k$ of $u_j$ is also in $I$.

@@

Now, 
fix an $R$-module $M$ generated by $u_1, \ldots, u_{n+1} \in M$. 
By assumption, \(M = \mathfrak{m} M\), and in particular, 
the generator $u_1$ is an element of the submodule $\mathfrak{m}M$.
The lemma described above guarantees the existence of ring elements $a_1, \ldots, a_n \in \mathfrak{m}$
for which 
\[u_{1} = a_{1} \cdot u_{1} + a_2 \cdot u_2 + \cdots + a_{n+1} \cdot u_{n+1}. \]

Collecting the \(u_{1}\) terms
on the left-hand side of the equation, applying the existence of additive inverses in $R$, and
 distributivity of scalar multiplication for modules, we see that
\begin{equation} \label{equation1: e}
(1 - a_{1}) \cdot u_{1} = a_{2} \cdot u_{2} + \cdots + a_{n+1} \cdot u_{n+1}\text{.}
\end{equation}

Since a proper ideal contains no units and $a_1$ is taken to be in $\mathfrak{m}$, it is not a unit.  
Hence, by the formalized statement mentioned earlier, $1-a_1$ is a unit of $R$.    
Let $b_{1} \in R$ denote its 
multiplicative inverse, so that $b_1 \cdot  (1 - a_{1}) = 1$.
Multiplying this element on the either side of \eqref{equation1: e}, we deduce the following.  
\begin{align*}
b_1 \cdot \left(  (1 - a_{1}) \cdot u_{1} \right)  & = b_1 \cdot  ( a_{2} \cdot u_{2} + \cdots + a_{n+1} \cdot u_{n+1} ) \\
\left( b_1 \cdot  (1 - a_{1})\right) \cdot u_{1}   & = b_1 \cdot  ( a_{2} \cdot u_{2} ) + \cdots + b_1  \cdot   (a_{n+1} \cdot u_{n+1} )  \\
u_1 = 1 \cdot u_{1}   & = ( b_1 \cdot  a_{2}  ) \cdot u_{2}  + \cdots + ( b_1  \cdot  a_{n+1} ) \cdot u_{n+1}  
\end{align*}
In particular, the generator $u_1$ can be written as a scalar combination of the $n$ generators $u_2, \ldots, u_{n+1}$, and is superfluous. Hence $M$ can be generated by $n$ elements, and $M=0$ by the inductive hypothesis.

\emily[inline]{Emily:  Reflecting, I think it would be valuable to include at least parts of the formal proof.  Pieces would work, though it is short, and it could be ``broken up'' among the description of the informal argument.  As is, our description of the proof focuses on the informal argument, which is not ``new''. }

@@

>>>>

The importance of quotients in commutative algebra motivates our use of
equivalence relations to define the components of a group structure. If one
were to instead use the regular Leibniz equality, it would be difficult to
identify a quotient group and another group. However, by  defining a group in
terms of an arbitrary equivalence relation, in our theory a quotient group is 
simply defined as a group, but under an equivalence relation that is not the
usual equality. Not much is lost due to Coq's rewrite tactics for
setoids--types equipped with an equivalence relation---can still be called
upon.

@@

This argument has potentially infinitely many steps. A proof with infinitely
many steps is a problem in Coq. To see this from the above argument, suppose we
wrote a proof tactic \mintinline{coq}{generate_larger_ideal} that allows us to
go from one non-maximal ideal \(I_{k}\) which contains \(x\) to another ideal
\(I_{k+1}\) which also contains \(x\). Provided our first ideal \(I_{1}\) is
not maximal, our first invocation of \mintinline{coq}{generate_larger_ideal}
will produce another ideal \(I_{2}\). Our new ideal \(I_{2}\) may also not be
maximal, requiring that we invoke \mintinline{coq}{generate_larger_ideal}
again. This can lead to an infinite number of calls to the tactic
\mintinline{coq}{generate_larger_ideal} which will occupy any machine's
processor(s) forever, and further, the infinite number of ideals in the proof
state will fill any machine's memory. After we have overwhelmed any system with
these infinite number of calls to a single proof tactic, we still have to
invoke more tactics to finish the proof with an application of Zorn's lemma,
which finally guarantees us a maximal ideal containing \(x\). To summarize, we
are forced to run the following proof script which will never get past calling
\mintinline{coq}{generate_larger_ideal} infinitely many times.
\begin{minted}{coq}
  repeat generate_larger_ideal. (* must be called infinitely often *)
  zorns_lemma. (* called after the above non-terminating call finishes *)
\end{minted}
We chose to avoid this infinite proof script issue by including an axiom that
any non-unit \(x\) of a ring must be contained in some maximal ideal.

@@

\begin{minted}{coq}
Context {P_ideal: Ideal Radd Rzero Rminus Rmul P}.
Context {P_maxideal: maximal_ideal Requiv Radd Rzero Rminus Rmul P}.
Context {R_local: local_ring Requiv Radd Rzero Rminus Rmul}.

Let ideal_module_pred := ideal_module Mequiv Madd Mzero action P.

Theorem nakayama:
  forall {n: nat}(basis: t M n),
    finitely_generated Mequiv Madd Mzero action basis ->
  (forall a: M, ideal_module_pred a) ->
  forall a: M, a =M= Mzero.
\end{minted}

To use as a building block in the formal proof of Nakayama's Lemma, we first
formally stated and proved a lemma that gives a concrete description of the
elements of the submodule $\mathfrak{m} M$  appearing in the statement of the
theorem. This lemma applies more generally, to any finitely generated module
$M$ over a (not necessarily local) commutative ring, and any submodule of the
form $I M$, for $I$ an arbitrary ideal. By definition, $IM$ consists of scalar
combinations of elements of $M$ with coefficients in $I$. The lemma states that
the elements from $M$ appearing in the expression can be chosen to be from any
fixed finite generating set for $M$. In other words, given any generating set
$u_{1},\dots, u_{n} \in M$  of a  finitely generated $R$-module $M$, every
element $x$ of $I M$ can be written as a scalar combination $a_1 \cdot u_1 +
a_2  \cdot u_2 + \cdots + a_n  \cdot u_n$, where each $a_i \in I$.

@@

The proof follows from a straightforward argument based on definitions, by
induction on the number of elements in a fixed generating set for $M$, which we
informally describe: By definition, $x \in I M$ can  be written, for some
positive integer $k$ and elements $r_i \in R$ and $w_i \in M$,
$1 \leq i \leq k$, as $x = r_1 \cdot w_1 + r_2 \cdot w_2 + \cdots +
r_k \cdot w_k$. What's more, by definition of a finite generating set $u_{1},
\dots, u_{n}$ for $M$, each $w_i$ equals $c_{i1} \cdot u_1 + c_{i2} \cdot u_2 +
\cdots + c_{in} \cdot u_n$ for appropriate choices of $c_{ij} \in R$. Hence,
inductively applying associativity, 
\begin{align*}
  x = \sum_{i=1}^{k} r_i \cdot w_i
    = \sum_{i=1}^{k} r_i \cdot \left(\sum_{j=1}^n c_{ij} \cdot  u_j\right)
    = \sum_{j=1}^{n}\sum_{i=1}^{k} r_i \cdot (c_{ij} \cdot u_j) 
    = \sum_{j=1}^{n}\left(\sum_{i=1}^{k} c_{ij} \cdot r_i\right) \cdot u_j.   
\end{align*}
By absorption in ideals, each  $c_{ij}  \cdot r_i$ is in $I$, and since ideals
are closed under addition, we inductively conclude that the coefficient $a_j
:= c_{1j} \cdot  r_1 +  c_{2j} \cdot  r_2 +  \cdots c_{kj}\cdot r_k$ of $u_j$
is also in $I$.

@@

Now, fix an $R$-module $M$ generated by $u_1, \ldots, u_{n+1} \in M$. By
assumption, \(M = \mathfrak{m} M\), and in particular, the generator $u_1$ is
an element of the submodule $\mathfrak{m} M$. The lemma described above
guarantees the existence of ring elements $a_1, \ldots, a_n \in \mathfrak{m}$
for which 
\[u_{1} = a_{1} \cdot u_{1} + a_2 \cdot u_2 + \cdots + a_{n+1} \cdot u_{n+1}.\]

Collecting the \(u_{1}\) terms on the left-hand side of the equation, applying
the existence of additive inverses in $R$, and distributivity of scalar
multiplication for modules, we see that
\begin{equation}\label{equation1: e}
(1 - a_{1}) \cdot u_{1} =
  a_{2} \cdot u_{2} + \cdots + a_{n+1} \cdot u_{n+1}\text{.}
\end{equation}
In Coq, this bit looks like the following code below.
\begin{minted}[linenos,firstnumber=157,stepnumber=5]{coq}
  assert ((Rone [+] Rminus x1) <.> u1 =M=
  linear_combin Madd Mzero action coeffs' generatingSet').
  { setoid_rewrite (module_distrib_Radd Radd Rmul Rone
      Mequiv Madd Mzero Mminus action).
    setoid_rewrite (module_Rone Radd Rmul Rone
      Mequiv Madd Mzero Mminus action).
    setoid_rewrite (module_minus_l Requiv Radd Rzero Rminus Rmul Rone
      Mequiv Madd Mzero Mminus action).
    symmetry.
    apply (group_move_r Mequiv Madd Mzero Mminus).
    symmetry.
    assumption. }
\end{minted}

Since a proper ideal contains no units and $a_1$ is taken to be in
$\mathfrak{m}$, it is not a unit. Hence, by the formalized statement mentioned
earlier, $1 - a_{1}$ is a unit of $R$. Let $b_{1} \in R$ denote its 
multiplicative inverse, so that $b_1 \cdot  (1 - a_{1}) = 1$. Multiplying this
element on the either side of \eqref{equation1: e}, we deduce the following.  
\begin{align*}
b_1 \cdot \left((1 - a_{1}) \cdot u_{1} \right)
  & = b_1 \cdot (a_{2} \cdot u_{2} + \cdots + a_{n+1} \cdot u_{n+1}) \\
\left( b_1 \cdot (1 - a_{1})\right) \cdot u_{1}
  & = b_1 \cdot (a_{2} \cdot u_{2}) + \cdots + b_1 \cdot (a_{n+1} \cdot u_{n+1}) \\
u_1 = 1 \cdot u_{1}
  & = (b_1 \cdot  a_{2}) \cdot u_{2} + \cdots + (b_1  \cdot  a_{n+1}) \cdot u_{n+1}  
\end{align*}
\begin{minted}[linenos,firstnumber=181,stepnumber=5]{coq}
  pose proof
    (local_comm_ring_sub_1_nonunit Requiv Radd Rzero Rminus Rmul Rone
      R_local x1 Hx1_nonunit) as H1mx1_unit.
  (* u1 = (b1 coeffs') . generatingSet *)
  inversion_clear H1mx1_unit as [b1 Hb1].
  setoid_rewrite (commutative Requiv Rmul) in Hb1.
  apply (module_op_l Requiv Mequiv action) with (r:=b1) in H.
  setoid_rewrite <- (module_distrib_Rmul Radd Rmul Rone
    Mequiv Madd Mzero Mminus action) in H.
  setoid_rewrite Hb1 in H.
  setoid_rewrite (module_Rone Radd Rmul Rone
    Mequiv Madd Mzero Mminus action) in H.
  setoid_rewrite <- (module_linear_combin_mul_l Requiv Radd Rmul Rone
    Mequiv Madd Mzero Mminus action) in H.
\end{minted}
In particular, the generator $u_1$ can be written as a scalar combination of
the $n$ generators $u_2, \ldots, u_{n+1}$, and is superfluous. Hence $M$ can be
generated by $n$ elements, and $M=0$ by the inductive hypothesis.

\emily[inline]{Emily: Reflecting, I think it would be valuable to include at
least parts of the formal proof.  Pieces would work, though it is short, and it
could be ``broken up'' among the description of the informal argument.  As is,
our description of the proof focuses on the informal argument, which is not
``new''.}

@@

\section{Conclusion}
TODO
